Verifying Spatial Properties of Array Computations

### 2017-04-14 oopsla17/study.tex
Typo
first results for hypothesis 1

\begin{tabular}{l|l|l}
  Shorthand & Classification (of subscripts) & Example \\ \hline
 (\textsf{vars}) & Variables (just left-hand side) & \fortran{x = ...} \\
 (\textsf{const}) & Constants & \fortran{a(0, 1)} \\
 (\textsf{IVs}) & Induction variables & \fortran{a(i, j)} \\
 (\textsf{neigh}) & Neighbourhood offsets (of the form
                                                  $i \pm c$) & \fortran{a(i, j-1)} \\
 (\textsf{neigh+c}) & Neighbourhood offsets and constants &
                                                            \fortran{b(i, 0, j+1)} \\
 (\textsf{aff}) & Affine offsets (of the form $a * i \pm b$) & 
                                     \fortran{x(2*i+1,j)} \\
 (\textsf{aff+c}) & Affine offsets and constants & \fortran{x(i+1, 0, 3*j+2)}
\end{tabular} \\[1em]

\noindent
The classification is computed from all information that flows into an
assignment, for example, the following decomposes a one-dimensional
three-point pattern across multiple intermediate assignments:
%%
\begin{minted}{fortran}
do i = 1, n
  x = a(i)
  y = a(i+1)
  b(i) = (a(i-1) + x + y)/3.0
end do
\end{minted}
%%x
Our analysis recognises this as one array computation (rather than three), starting at
line 4, and reading from subscripts \fortran{a(i)}, \fortran{a(i+1)},
\fortran{a(i-1)}. We thus traverse the body of a loop in reverse,
using reaching-definitions to calculate of the array reads that flow
to a particular assignment. Since \fortran{x} and \fortran{y} flow to
the line 4, line 2 and line 3 are not classified, though multiple
assignments can flow to multiple classified statements.

@@

\noindent
Finally, we categorised the relationship between the left-hand side
and right-hand side, in terms of the use of induction variables.
The two sides are \emph{consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the left-hand side
induction variables are either a 
a subset or superset of the induction variables on the right-hand
side. Otherwise, the two sides are seen as inconsistent:

@@

\paragraph{Addressing the hypotheses}

\begin{enumerate}
\item Loops over arrays mainly read from arrays with a static pattern
based on constant offsets from (base or dervied) induction variables;
\end{enumerate}

@@

>>>>

\begin{tabular}{l|l|l}
  Shorthand & Classification (of subscripts) & Example \\ \hline
 (\textsf{vars}) & Variables (just LHS) & \fortran{x = ...}
  \\
 (\textsf{IVs}) & Induction variables (just LHS) & \fortran{a(i, j)} \\
 (\textsf{const}) & Constants & \fortran{a(0, 1)} \\
 (\textsf{neigh}) & Neighbourhood offsets (of the form
                                                  $i \pm c$) & \fortran{a(i, j-1)} \\
 (\textsf{neigh+c}) & Neighbourhood offsets and constants &
                                                            \fortran{b(i, 0, j+1)} \\
 (\textsf{aff}) & Affine offsets (of the form $a * i \pm b$) &
                                     \fortran{a(2*i+1,j)} \\
 (\textsf{aff+c}) & Affine offsets and constants & \fortran{a(i+1, 0,
                                                   3*j+2)} \\
 (\textsf{other}) & Subscript expression not included above &
\fortran{x(f(i))}, \fortran{a(i `mod` 2)}
\end{tabular} \\[1em]

\noindent
Note that the \textsf{vars} and \textsf{IVs} are only give to
left-hand sides.
The classification is computed from all information that flows into an
assignment, for example, the following decomposes a one-dimensional
three-point pattern across multiple intermediate assignments:
%%
\begin{minted}{fortran}
do i = 1, n
  x = a(i)
  y = a(i+1)
  b(i) = (a(i-1) + x + y)/3.0
end do
\end{minted}
%%x
Our analysis recognises this as one array computation (rather than three), starting at
line 4, and reading from subscripts \fortran{a(i)}, \fortran{a(i+1)},
\fortran{a(i-1)}. We thus traverse the body of a loop in reverse,
using reaching-definitions to calculate of the array reads that flow
to a particular assignment. Since \fortran{x} and \fortran{y} flow to
the line 4, line 2 and line 3 are not classified, though multiple
assignments can flow to multiple classified statements.

@@

\noindent
Finally, we categorised the relationship between the left-hand side
and right-hand side, in terms of the use of induction variables.
The two sides are \emph{consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the left-hand side
induction variables are either
a subset or superset of the induction variables on the right-hand
side. Otherwise, the two sides are seen as inconsistent:

@@

%\begin{enumerate}
%\item Loops over arrays mainly read from arrays with a static pattern
%based on constant offsets from (base or dervied) induction variables;
%\end{enumerate}

@@

\noindent
We identified $108,773$ instances of assignments within loops in
which an array subscript flows to the right-hand side. We refer
to each one of these as an \emph{array computation}.
We performed the classification described in the methodology,
and then grouped the data by based only on the right-hand side
classification:
%%
\begin{center}
\begin{tabular}{lrrcc}
\textbf{RHS classification} & \textbf{Number} & \textbf{\%} &
\textit{grouping 1} & \textit{grouping 2} \\ \hline
Affine                          & 34        & 0.03\%  &
                                                            \multirow{4}{*}{92.61\%}
  \\ \cline{1-3}\cline{5-5}
All constant                    & 12160     & 11.18\%  \\ \cline{1-3}\cline{5-5}
Neighbourhood                   & 65334     & 60.06\% &
                                                          & \multirow{2}{*}{81.4\%}
                                                           \\ \cline{1-3}
Neighbourhood + constants       & 23209     & 21.34\%  \\ \hline
Other                           & 8036      & 7.39\%  \\ \hline \hline
Total                           & 108773    &  \\
\end{tabular}
\end{center}

\noindent
Since, affine, consant, and neighbour are all static patterns, we
group them (\textit{grouping 1}) to see that $92.61\%$ of the array computations have
a static pattern that are either constants or constant of offsets
of base/derived induction variables, thus confirming
\emph{hypothesis 1}. Note that affine (derived induction variables) are very
rare. Overwhelmingly the main category is neighbourhood offsets
(\emph{grouping 2})
(possibly mixed with constants in some dimensions), at $81.4\%$ in
total. The ``affine + constants'' class is not represented at all.
